---
title: "Introduction to ALKfusiondiscover"
output: rmarkdown::html_vignette
date: '2022-07-18'
vignette: >
  %\VignetteIndexEntry{Introduction to ALKfusiondiscover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

This is an introduction to the **ALKfusiondiscover** package which can be used to evaluate whether EML4-ALK is present in blood plasma. 

This package was created in order to increase the sensitivity of EML4-ALK detection from commercially available NGS products such the AVENIO (Roche) pipeline. 

Paired-end sequencing of cfDNA generated BAM files can be used as input to
discover EML4-ALK variants. This package was developed using position deduplicated BAM files generated with the AVENIO Oncology Analysis Software. These files are made using the AVENIO ctDNA surveillance kit and Illumina Nextseq 500 sequencing. This is a targeted hybridization NGS approach and includes ALK-specific but not EML4-specific probes.

The package includes six functions.

The output of first function, `EML4_ALK_detection()`, is used to determine whether EML4-ALK is detected and serves as input for the next four exploratory functions characterizing the EML4-ALK variant. The last function `EML4_ALK_analysis()` combines the output of the exploratory functions. 

To serve as examples, this package includes BAM files representing the EML4-ALK positive cell line H3122 and the EML4-ALK negative cell line, HCC827.  

## Installation

Use **devtools** to install **ALKfusiondiscover**.

```{r setup, message=FALSE}
if (!require(devtools)) install.packages('devtools')
library(devtools)

install_github("CTrierMaansson/ALKfusiondiscover")
library(ALKfusiondiscover)

```

## Package data

BAM files from the cell lines, H3122 and HCC827, are included in the package and can be used as examples to explore the functions. 

```{r examples}
H3122_bam <- system.file("extdata", "H3122_EML4.bam", package = "ALKfusiondiscover")
HCC827_bam <-  system.file("extdata", "HCC827_EML4.bam", package = "ALKfusiondiscover")

```

## EML4_ALK_detection


